Visualizing Contingency Tables

### 2007-10-09 trunk/vcd/paper/Handbook/hsc.Rnw
* enhancement of legend (min/max in range can be set) * fixes from proof-reading added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@659 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<color,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e., appear identical.}
\label{fig:colors}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
plot.new()
rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5, c = 100, l = c(50, 90)))
rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
@ 
\caption{Diverging color palettes for the HSV space (upper part) and
  the HCL space (lower part), ranging from blue over a neutral color to
  red. The triples indicate the settings for 
  the three dimensions: (Hue, Saturation, Value)
  in the upper part, and (Hue, Chroma, Luminance) in the lower part.
  In a monochrome version of the paper, the right and left hand
  sides of the HCL colors palette appear identical. }
\label{fig:diverging}
\end{center}
\end{figure}

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
\index{color space!CIELAB} \index{color space!CIELUV}
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
\index{color space!HCL}
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Figure~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Figure~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminance better 
contrasts can be achieved.

@@

In a first step, we create a mosaic plot for an exploratory view of the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used as the last split via highlighting. 
Furthermore, we expect age and education to have an influence on both memory
and attitude, hence they should be used first for splitting as stratifying
variables. In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people are more than 40 years of
age, most of whom of only completed elementary school (much more than 
in the other age groups). Furthermore, it can be seen that the proportion
of people with memories of punishments increases with age and that
the approval rate decreases with education (this, however, is better shown in 
Figure~\ref{fig:punishcotab} described below). As for the question whether
attitude depends on memory, the plot suggests quite clearly (and somewhat
surprisingly) that for those people with elementary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without: 
experience of punishment seems to support use of punishment.
For the other education groups, the picture is less clear: some cells 
indicate the same association, while others do not.

@@

Mosaic, association, and sieve plots can be used for visualizing multi-way tables
by applying them to flat representations. Mosaic plots are particularly useful
for exploratory analysis, whereas the other two require specification of a certain
independence model from which deviations should be examined. In addition, specialized
`flavours' of the plots can leverage
exploratory analyses (pairs plots, highlighting, doubledecker plots) 
or model-based analyses (residual-based shading, conditional plots).

>>>>

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<color,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@
\caption{Qualitative color palette for the HSV (left)
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e., they will
         appear to be virtually identical.}
\label{fig:colors}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
plot.new()
rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5, c = 100, l = c(50, 90)))
rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
@
\caption{Diverging color palettes for the HSV space (upper part) and
  the HCL space (lower part), ranging from blue over a neutral color to
  red. The triples indicate the settings for
  the three dimensions: (Hue, Saturation, Value)
  in the upper part, and (Hue, Chroma, Luminance) in the lower part.
  In a monochrome version of the paper, the right and left hand
  sides of the HCL colors palette appear to be identical, unlike the
  HSV color palette. }
\label{fig:diverging}
\end{center}
\end{figure}

The use of colors that are more `in harmony' goes back to Munsell
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
\index{color space!CIELAB} \index{color space!CIELUV}
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
\index{color space!HCL}
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance
are always balanced towards the same grey
and thus do not have the problem of varying saturations as the HSV
colors (see Figure~\ref{fig:colors}, right).
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2007} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Figure~\ref{fig:diverging}.
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue)
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter
full chroma colors), however, by varying both chroma and luminance better
contrasts can be achieved.

@@

In a first step, we create a mosaic plot for an exploratory view of the data:
attitude towards punishment is clearly the response variable
here and should be used as the last split via highlighting.
Furthermore, we expect age and education to have an influence on both memory
and attitude, hence they should be used first for splitting as stratifying
variables. In fact, the question is whether memory has some influence on attitude, given
age and education. We choose to create a three-way mosaic plot of the
hypothesized explanatory variables (first splitting horizontically by age,
then vertically by education, and finally horizontically again by memory),
and to have attitude, the response variable,
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people are more than 40 years of
age, most of whom of only completed elementary school (much more than
in the other age groups). Furthermore, it can be seen that the proportion
of people who recall being punished increases with age, and that
the approval rate decreases with education (this, however, is better shown in
Figure~\ref{fig:punishcotab} described below). As for the question whether
attitude depends on memory, the plot suggests quite clearly (and somewhat
surprisingly) that for those people with elementary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without:
experience of punishment seems to support use of punishment.
For the other education groups, the picture is less clear: some cells
indicate the same association, while others do not.

@@

Mosaic, association, and sieve plots can be used for visualizing multi-way tables
by applying them to flat representations. Mosaic plots are particularly useful
for exploratory analysis, whereas the other two require specification of a certain
independence model from which deviations should be examined. In addition, specialized
`flavours' of the plots can leverage
exploratory analyses (pairs plots, highlighting, doubledecker plots)
or model-based analyses (residual-based shadings, conditional plots).
